The Leaky Semicolon (Supplementary Material)

### 2021-11-10 paper/abstract.tex
edit sections 1-3

>>>>

Program logics and semantics tell a pleasant story about sequential composition: when executing (S1;S2), we first execute S1 then S2. To improve performance, however, processors execute instructions out of order, and compilers reorder programs even more dramatically. By design, single-threaded systems cannot observe these reorderings; however, multiple-threaded systems can, making the story considerably less pleasant. A formal attempt to understand the resulting mess is known as a “relaxed memory model.” Prior models either fail to address sequential composition directly, or overly restrict processors and compilers, or permit nonsense thin-air behaviors which are unobservable in practice.

To support sequential composition while targeting modern hardware, we enrich the standard event-based approach with preconditions and families of predicate transformers. When calculating the meaning of (S1; S2), the predicate transformer applied to the precondition of an event e from S2 is chosen based on the set of events in S1 upon which e depends. We apply this approach to two existing memory models.

### 2021-11-10 paper/intro.tex
edit sections 1-3

These models are all non-compositional in the sense that in order to
calculate the meaning of any thread, all threads must be known.  Using the
axiomatic approach of \cXI{}, for example, execution graphs are first
constructed for each thread, using an operational semantics that allows a
read to see any value.  The combined graphs are then filtered using a set of
acyclicity axioms that determine which reads are valid.  These axioms use
existentially defined global relations, such as memory order $\rmo$, which
must be a per-location total order on write actions.

@@

Our main insight is that for language models, \emph{sequentiality} is the
hard part.  \emph{Concurrency} is easy!  Or at least, it is no more difficult
than it is for hardware.  Compilers make the difference, since they typically
do little optimization between threads.  We motivate our approach to
sequential dependencies in \textsection\ref{sec:overview} and provide formal
definitions in \textsection\ref{sec:model}.  We extend the model to include
additional features, such as address calculation and \RMW{}s, in
\textsection\ref{sec:additional}.  We discuss related and future work in
\textsection\ref{sec:related}--\ref{sec:outro}.

@@

For a sequentially consistent semantics, sequential composition is
straightforward: we apply each predicate transformer to subsequent
preconditions, composing the predicate transformers.
\begin{equation}
\begin{gathered}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{rx0}{\DR{x}{1}}{}
      \event{ry0}{\DR{y}{1}}{right=of rx0}
      \event{wz0}{(1{=}r) \limplies (1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm[r{*}(s{-}1)/z]}{right=of wz0}
      \wki{rx0}{ry0}
      \wki{ry0}{wz0}
      %\xo{rx0}{rx0ry0d}
      %\xo[out=155,in=15]{ry0}{rx0ry0d}
      % \xo{wz0}{rx0ry0d}
  \end{tikzinline}}
\end{gathered}
\label{eq1}
\end{equation}
This works for the sequentially consistent case, but needs to be
weakened for the relaxed case.

@@

For example, we can recover the expected semantics
for \eqref{eq1} by choosing
the predicate transformer which is independent of $(\DR x1)$
but dependent on $(\DR y1)$, which is the transformer
which maps $\bForm$ to $(1{=}s) \limplies \bForm$.
(In subsequent diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=2ex and 1.2em]
      \xform{wz0i}{\bForm}{}
      \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of wz0i}
      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm}{right=of rx0d}
      \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of rx0ry0d}
      \event{rx0}{\DR{x}{1}}{above left=of rx0ry0d}
      \event{ry0}{\DR{y}{1}}{above right=of rx0ry0d}
      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
As a quick correctness check, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left---with the intermediate step as in the diagram above,
eliding the write action---
% \begin{gather*}
%   \PR{x}{r}\SEMI \PR{y}{s}
%   \\
%   \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
%       \xform{rx0d}{(0{=}r) \limplies \bForm}{}
%       \event{rx0}{\DR{x}{0}}{right=of rx0d}
%       \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
%       \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
%       \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
%       \xform{wz0i}{\bForm}{left=of rx0d}
%       \xo{rx0}{rx0d}
%       \xo{ry0}{ry0d}
%       \xo{rx0}{rx0ry0d}
%       \xo{ry0}{rx0ry0d}
%     \end{tikzinline}}
% \end{gather*}
or to the right---with the intermediate step:
\begin{gather*}
  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{ry0}{\DR{y}{1}}{}
      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{ry0d}{(1{=}s) \limplies \bForm}{left=of ry0}
      \xform{wz0i}{\bForm}{left=of ry0d}
      \po{ry0}{wz0}
      \xo{ry0}{ry0d}
    \end{tikzinline}}
\end{gather*}
This is an instance of the general result that sequential composition forms a
monoid (\reflem{lem:monoid}).

>>>>

These models are all non-compositional in the sense that in order to
calculate the meaning of any thread, all threads must be known.  Using the
axiomatic approach of \cXI{}, for example, execution graphs are first
constructed for each thread, using an operational semantics that allows a
read to see any value.  The combined graphs are then filtered using a set of
acyclicity axioms that determine which reads are valid.  These axioms use
existentially defined global relations, such as memory order ($\rmo$), which
must be a per-location total order on write actions.

@@

Our main insight is that for language models, \emph{sequentiality} is the
hard part.  \emph{Concurrency} is easy!  Or at least, it is no more difficult
than it is for hardware.  Compilers make the difference, since they typically
do little optimization between threads.  We motivate our approach to
sequential dependencies in \textsection\ref{sec:overview} and provide formal
definitions in \textsection\ref{sec:model}.  In
\textsection\ref{sec:additional}, we extend the model to include
additional features, such as address calculation and \RMW{}s.  We discuss related and future work in
\textsection\ref{sec:related}--\ref{sec:outro}.

@@

For a sequentially consistent semantics, sequential composition is
straightforward: we apply each predicate transformer to subsequent
preconditions, composing the predicate transformers.
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \tag{*}\label{eq1}
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{rx0}{\DR{x}{1}}{}
      \event{ry0}{\DR{y}{1}}{right=of rx0}
      \event{wz0}{(1{=}r) \limplies (1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm[r{*}(s{-}1)/z]}{right=of wz0}
      \wki{rx0}{ry0}
      \wki{ry0}{wz0}
      %\xo{rx0}{rx0ry0d}
      %\xo[out=155,in=15]{ry0}{rx0ry0d}
      % \xo{wz0}{rx0ry0d}
  \end{tikzinline}}
\end{gather*}
This works for the sequentially consistent case, but needs to be
weakened for the relaxed case.

@@

For example, we can recover the expected semantics
for \eqref{eq1} by choosing
the predicate transformer which is independent of $(\DR x1)$
but dependent on $(\DR y1)$, which is the transformer
which maps $\bForm$ to $(1{=}s) \limplies \bForm$.
(In subsequent diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=2ex and 1.2em]
      \xform{wz0i}{\bForm}{}
      \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of wz0i}
      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm}{right=of rx0d}
      \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of rx0ry0d}
      \event{rx0}{\DR{x}{1}}{above left=of rx0ry0d}
      \event{ry0}{\DR{y}{1}}{above right=of rx0ry0d}
      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
As a quick correctness check, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left---with the intermediate step as in the diagram above,
eliding the write action---or to the right---with the intermediate step:
% \begin{gather*}
%   \PR{x}{r}\SEMI \PR{y}{s}
%   \\
%   \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
%       \xform{rx0d}{(0{=}r) \limplies \bForm}{}
%       \event{rx0}{\DR{x}{0}}{right=of rx0d}
%       \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
%       \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
%       \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
%       \xform{wz0i}{\bForm}{left=of rx0d}
%       \xo{rx0}{rx0d}
%       \xo{ry0}{ry0d}
%       \xo{rx0}{rx0ry0d}
%       \xo{ry0}{rx0ry0d}
%     \end{tikzinline}}
% \end{gather*}
\begin{gather*}
  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{ry0}{\DR{y}{1}}{}
      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{ry0d}{(1{=}s) \limplies \bForm}{left=of ry0}
      \xform{wz0i}{\bForm}{left=of ry0d}
      \po{ry0}{wz0}
      \xo{ry0}{ry0d}
    \end{tikzinline}}
\end{gather*}
This is an instance of the general result that sequential composition forms a
monoid (\reflem{lem:monoid}).
